Astrophysical calibration of gravitational-wave detectors

### 2015-07-07 docs/cosmic_cal.tex
minor changes to the paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[10pt]{iopart}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{iopams}
\usepackage{color}

%\documentclass[12pt]{iopart}
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
%\usepackage {amsmath}
%\usepackage {amssymb}
%\usepackage {amsfonts}
%\usepackage {amsthm}
%\usepackage {mathrsfs}
%\usepackage {natbib}
%\usepackage {latexsym}
%\usepackage {graphicx}
%\usepackage {dsfont}
%\usepackage {times}
%\usepackage {txfonts}
%\usepackage {rotating}
%\usepackage {wasysym}
%\usepackage {multirow}
%\usepackage {hhline}
\usepackage{hyperref}
%\usepackage {color}
\usepackage{bm}
%\usepackage{appendix}
\usepackage{acronym}
%\usepackage{dcolumn}   % needed for some tables
\usepackage{url}
\usepackage[normalem]{ulem}   % temporal one in draft.
\usepackage{xspace}

@@

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{NSBH}[NSBH]{neutron star black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}

@@

% Describe our model for the calibration
\MP{The gravitational wave strain is measured through the differential arm length, $\Delta L$, changes
of the interferometer via $h = \Delta L(f) / L$, where $L$ is the full arm length.
Calibration is required to relate the actual measured interferometer error signal output
$e(f)$ to $\Delta L$. This relation is known as the length response function, $R(f,t)$,
defined such that
\begin{equation}
\Delta L(f) = R(f,t) e(f),
\end{equation}
where $R$ varies slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to
uncertainties. In this study we will assume an estimate of $R$ is available (although
in theory we could take on the role of estimating $R$ itself), but that it differs from the
truth through some unknown scale factor, $s$, so that
\begin{equation}
h(f) = s h_{\rm m}(f) = s \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain.
In this analysis we will simplify the situation by assuming that $s$ is a constant with
respect to frequency, but in future studies that assumption could be dropped and $s$
could take some functional form, or piece-wise fit, with respect to $f$.}

@@

% Describe what we are going to do in this section
Let us consider a single \ac{GW} detection of the inspiral stage of a \ac{BNS}
merger and assume that this signal has been associated with a potential
\ac{sGRB} detection. We therefore also assume that the inclination angle
of the binary is consistent with a given \ac{sGRB} beaming angle.  Finally we
assume that a host galaxy has been identified for the source and that the sky
position is therefore known to high precision.

\MP{We perform 100 simulations at each of a range of distances from 10\,Mpc out
to 200\,Mpc. For the simulations at each distance range we have allowed the
source's sky position to be randomly drawn from a uniform distribution over
the sky, the polarisation angle $\psi$ to drawn from a uniform distribution in the
range $[0, \pi/2]$, the coalescence phase to be drawn from a uniform distribution
in the range $[0, 2\pi]$, and the inclination angle drawn from a Gaussian with range XX?.
The binary system masses have been fixed as $1.4\,{\rm M}_\odot$ (based on the
very good assumption that whatever masses are chosen in these cases they will be
recovered to high accuracy) and the time of coalescence is fixed.}

% Describe the signal model (Taylor F2 etc...)
We use the TaylorF2 waveform approximation (see e.g.\ \cite{2009PhRvD..80h4043B} and references 
therein) with a 3.5 post-Newtonian expansion in phase for modelling our signal (in both simulations 
and signal recovery).  For \ac{BNS} systems we use non-spinning waveforms under the assumption that 
spins will be negligible. For \ac{NS-BH} system we use a spinning waveform, but in which only the 
black hole has non-negligible spin and that its rotational angular momentum is aligned with the 
systems orbital angular momentum.

@@

\MP{Our aim is to estimate the marginal posterior probability distribution function (pdf)
for the unknown calibration scale factor. We use Bayes' theorem for which we need to define
a likelihood function and prior pdfs on the signal parameters being estimated. In this
analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to sample 
the posterior distribution.} We use a Gaussian likelihood function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
\fl p(\bi{d} | \btheta, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\Delta f$ is the frequency resolution of the data, $S_k$ is the $k^{\rm th}$ detector's one 
sided noise power spectral density, and the $i$ indices increment over frequency with $i_{f_{\rm 
low}}$ and $i_{f_{\rm high}}$ corresponding to the lower and upper range in frequencies used for 
our analysis, which were 30\,Hz to 1600\,Hz respectively. The source parameters, and the 
unknown scale factor, have been subsumed into a vector $\btheta$. For a network of detectors, 
the overall likelihood function becomes a product of the likelihoods for each detectors, i.e.\
\begin{equation}
  \label{eq:mult-likeli}
  p(\bi{d}| \btheta, \curlH) = \prod \limits_D p(\bi{d}^D| \vec{\theta}, \curlH)
\end{equation}
where $D = \{H,L,V\}$.

% describe what parameters we will be considering and their priors
We assume uniform prior distributions on the unknown nuisance parameters
$\phi_{\mathrm{c}},\psi$ on the ranges $[0, 2\pi]$ and $[0, \pi/2]$ respectively.
The time of coalescence is
restricted to a prior range of $\pm 10$ msec around the true value.  Given our
assumption of an \ac{sGRB} counterpart and host galaxy, we treat the sky
position as exactly known. We also consider the cosine of the inclination angle to be drawn
from a zero-mean Gaussian with standard deviation $0.1$ consistent with an
uncertainty on \ac{sGRB} opening of XXX~\cm{What was exactly done here?}.  The
main parameter of interest, the calibration scale factor $s$ is assumed to have
a uniform prior distribution on the range $(0,2)$.

@@

The primary reason why we can use coincident observations with a GRB as a potential check of 
detector calibration is that the GRB allows us to constrain our priors on various parameters that 
would normally be highly correlated with the signal amplitude. The most important of these is that 
the GRB can provide a very tight constraint on the source distance\footnote{A consequence of our 
choice of BNS signals is that they will only be observable within the local Universe ($z \sim 
0.1$), so the relation between the measured GRB redshift and distance is simple and is negligably 
effected by the choice of cosmology.}. In this analysis we therefore assume that the source distance 
is known (i.e.\ has a $\delta$-function prior). Another piece of information that we can use from 
having a coincident GRB is that the system must be close to face-on, which allows us to place prior 
constraints on the system inclination angle. In our analysis we take two scenarios: the first use a 
prior on the beam opening angle (half-width of the beam) with a Gaussian distribution with zero mean 
and $\sigma = 5^{\circ}$, whilst the second uses $\sigma = 20^{\circ}$. Beam opening angles for 
short GRBs are relatively poorly constrained as they are based on only a few jet break detection, 
however \cite{2014ApJ...780..118F} give a median opening angle value of $\sim 10^{\circ}$. If we 
assume that the beam opening angle is a half-normal distribution, for which the median is given by 
$\sqrt{2}\erf{}^{-1}(1/2)\sigma \sim 0.67\sigma$, then a median of $\sim 10^{\circ}$ corresponds to 
$\sigma \sim 14.8^{\circ}$.

@@

@@

To estimate how well we can constrain the calibration scaling each detector in an advanced 
detector network we have performed simulations with injected BNS signals at a range of distances 
(50\,Mpc to 450\,Mpc on 50\,Mpc increments). The simulations use the two Advanced LIGO detectors 
(one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced Virgo detector (V1). We 
assume all are expected to be operating at their design sensitivity.

>>>>

\documentclass[10pt]{iopart}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{iopams}
\usepackage{color}

%\usepackage {amsmath}
%\usepackage {amssymb}
%\usepackage {amsfonts}
%\usepackage {amsthm}
%\usepackage {mathrsfs}
%\usepackage {natbib}
%\usepackage {latexsym}
%\usepackage {graphicx}
%\usepackage {dsfont}
%\usepackage {times}
%\usepackage {txfonts}
%\usepackage {rotating}
%\usepackage {wasysym}
%\usepackage {multirow}
%\usepackage {hhline}
\usepackage{hyperref}
%\usepackage {color}
\usepackage{bm}
%\usepackage{appendix}
\usepackage{acronym}
%\usepackage{dcolumn}   % needed for some tables
\usepackage{url}
\usepackage[normalem]{ulem}   % temporal one in draft.
\usepackage{xspace}

@@

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}

@@

% Describe our model for the calibration
\MP{The gravitational wave strain is measured through the differential arm length, $\Delta L$, 
changes of the interferometer via $h = \Delta L(f) / L$, where $L$ is the full arm length.
Calibration is required to relate the actual measured interferometer error signal output
$e(f)$ to $\Delta L$. This relation is known as the length response function, $R(f,t)$,
defined such that
\begin{equation}
\Delta L(f) = R(f,t) e(f),
\end{equation}
where $R$ varies slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to
uncertainties. In this study we will assume an estimate of $R$ is available (although
in theory we could take on the role of estimating $R$ itself), but that it differs from the
truth through some unknown scale factor, $s$, so that
\begin{equation}
sh(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $s > 1$ then a signal would appear to have a larger amplitude (e.g.\ be closer) than 
expected, whereas if $s < 1$ would appear to have a lower amplitude than expected. In this analysis 
we will simplify the situation by assuming that $s$ is a constant with respect to frequency, but in 
future studies that assumption could be dropped and $s$ could take some functional form, or 
piece-wise fit, with respect to $f$.}

@@

% Describe the signal model (Taylor F2 etc...)
We use the TaylorF2 waveform approximation (see e.g.\ \cite{2009PhRvD..80h4043B} and references 
therein) with a 3.5 post-Newtonian expansion in phase for modelling our signal (in both simulations 
and signal recovery).  For \ac{BNS} systems we use non-spinning waveforms under the assumption that 
spins will be negligible. For \ac{NSBH} system we use a spinning waveform, but in which only the 
black hole has non-negligible spin and that its rotational angular momentum is aligned with the 
systems orbital angular momentum.

@@

\MP{Our aim is to estimate the marginal posterior probability distribution function (pdf)
for the unknown calibration scale factors. We use Bayes' theorem for which we need to define
a likelihood function and prior pdfs on the signal parameters being estimated. In this
analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to sample 
the posterior distribution.} We use a Gaussian likelihood function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
\fl p(\bi{d} | \btheta, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\Delta f$ is the frequency resolution of the data, $S_k$ is the $k^{\rm th}$ detector's one 
sided noise power spectral density, and the $i$ indices increment over frequency with $i_{f_{\rm 
low}}$ and $i_{f_{\rm high}}$ corresponding to the lower and upper range in frequencies used for 
our analysis, which were 20\,Hz to 400\,Hz respectively. This frequency range was chosen as the 
vast majority of the \ac{SNR} can be found within this range, so going to lower or higher 
frequencies provides very little additional information for our analysis. The source parameters, 
and the unknown scale factor, have been subsumed into a vector $\btheta$. For a network of 
detectors, the overall likelihood function becomes a product of the likelihoods for each detectors, 
i.e.\
\begin{equation}
  \label{eq:mult-likeli}
  p(\bi{d}| \btheta, \curlH) = \prod \limits_D p(\bi{d}^D| \vec{\theta}, \curlH)
\end{equation}
where $D = \{H,L,V\}$.

@@

The primary reason why we can use coincident observations with a GRB as a potential check of 
detector calibration is that the GRB allows us to constrain our priors on various parameters that 
would normally be highly correlated with the signal amplitude. The most important of these is that 
the GRB can provide a very tight constraint on the source distance\footnote{For \ac{BNS} and 
\ac{NSBH} systems they will be observable only in the relatively nearby Universe ($z \sim 
0.1$ and $z\sim 0.2$ for \ac{BNS} and \ac{NSBH} systems respectively), so the relation between the 
measured GRB redshift and distance is simple and our luminosity distance will be negligably 
effected by the choice of cosmology \MP{Check this?}.}. In this analysis we therefore assume that 
the source distance is known (i.e.\ has a $\delta$-function prior). Another piece of information 
that we can use from having a coincident GRB is that the system must be close to face-on, which 
allows us to place prior constraints on the system inclination angle. In our analysis we take two 
scenarios: the first use a 
prior on the beam opening angle (half-width of the beam) with a Gaussian distribution with zero mean 
and $\sigma = 5^{\circ}$, whilst the second uses $\sigma = 20^{\circ}$. Beam opening angles for 
short GRBs are relatively poorly constrained as they are based on only a few jet break detection, 
however \cite{2014ApJ...780..118F} give a median opening angle value of $\sim 10^{\circ}$. If we 
assume that the beam opening angle is a half-normal distribution, for which the median is given by 
$\sqrt{2}\erf{}^{-1}(1/2)\sigma \sim 0.67\sigma$, then a median of $\sim 10^{\circ}$ corresponds to 
$\sigma \sim 14.8^{\circ}$.

@@

For the spinning \ac{NSBH} systems we use a uniform prior on the dimensionless spin magnitude of 
the black hole between $-1$ and 1.

@@

To estimate how well we can constrain the calibration scaling each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 1000\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations use the 
two Advanced LIGO detectors (one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced 
Virgo detector (V1). The simulations are all performed in the freqeuncy domain and span a frequency 
range from 20\,Hz to 400\,Hz. We assume all detectors are expected to be operating at their design
sensitivity.
